Natural Selection Across Three Generations of Americans

### 2024-03-02 hrs-selection.qmd
Refined phenotypic change section

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: scores which correlate negatively (positively) with
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  partly borne out: coefficients are larger among low-income
  parents, but not among other groups. Using respondents' number of 
  siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

knitr::knit_hooks$set("inline", 
  \(x) {
    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
    }
    x
  }
)
nice_names <- c(
  SCZ = "Schizophrenia",
  WC = "Waist circ.",
  WHR = "Waist-hip ratio",
  NEUROT = "Neuroticism",
  WELLB = "Well-being",
  DEPSYMP = "Depr. symptoms",
  CD = "Cor. art. disease",
  MI = "Myoc. infarction",
  CORT = "Cortisol",
  T2D = "Type 2 diabetes",
  BIP = "Bipolar",
  ADHD_PGC17 = "ADHD",
  XDISORDER = "Cross-disorder",
  MENA = "Age at menarche",
  MENO = "Age at menopause",
  EXTRAVER = "Extraversion",
  AUTISM = "Autism",
  AB = "Antisoc. behav.",
  OCD = "OCD",
  AFBC = "Age 1st birth combn",
  AFBF = "Age 1st birth F",
  AFBM = "Age 1st birth M",
  MDD2 = "Maj. depr. disorder",
  PTSDAA = "PTSD Afr",
  PTSDEA = "PTSD Eur",
  PTSDC = "PTSD combined",
  HDL = "HDL cholesterol",
  LDL = "LDL cholesterol",
  TC = "Total cholesterol",
  ANXFS = "Anxiety factor",
  ANXCC = "Anxiety case-control",
  BUN = "Blood urea nitro.",
  BUNTE = "Blood urea nitro. TE",
  CKD = "Kidney disease",
  CKDTE = "Kidney disease TE",
  DBP = "Diastolic blood pr.",
  BMI2 = "BMI",
  HEIGHT2 = "Height",
  AI = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW = "Drinks per week",
  SC = "Smoking cessation",
  SI = "Smoking initiation",
  HTN = "Hypertension",
  CANNABIS = "Cannabis",
  GWAD2NA = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC = "Alcohol dependence",
  PP = "Pulse pressure",
  SBP = "Systolic blood pr.",
  EGFR = "Glom. filtr.",
  EGFRTE = "Glom. filtr. TE",
  EA3 = "Educ. attainment",
  HBA1CAA = "HbA1c Afr",
  HBA1CEA = "HbA1c Eur",
  GCOG2 = "Gen. cognition",
  ACTIVITY = "Physical Activity", # pgirep here
  ADHD = "ADHD",
  ADVENTURE = "Adventurous",
  AFB = "Age First Birth",
  ALLERGYCAT = "Cat Allergy",
  ALLERGYDUST = "Dust Allergy",
  ALLERGYPOLLEN = "Pollen Allergy",
  ASTECZRHI = "Asthma/Eczema/Rhinitis",
  ASTHMA = "Asthma",
  AUDIT = "Alcohol Misuse",
  BMI = "Body Mass Index",
  CANNABISpgirep = "Cannabis Use",
  COGEMP = "Cognitive Empathy",
  COPD = "Chronic Obstructive Pulmonary Disease",
  CPD = "Cigarettes per Day",
  CP = "Cognitive Performance",
  DELAYDISC = "Delay Discounting",
  DEP = "Depressive Symptoms",
  DPWpgirep = "Drinks per Week",
  EA = "Educational Attainment",
  EVERSMOKE = "Ever Smoker",
  EXTRA = "Exraversion",
  FAMSAT = "Satisfaction with Family",
  FINSAT = "Satisfaction with Finances",
  FRIENDSAT = "Satisfaction with Friends",
  HAYFEVER = "Hayfever",
  HEIGHTpgirep = "Height",
  HIGHMATH = "Highest Math",
  LEFTOUT = "Left Out of Social Activity",
  LONELY = "Loneliness",
  MENARCHE = "Age First Menses",
  MIGRAINE = "Migraine",
  MORNING = "Morning Person",
  NARCIS = "Narcissism",
  NEARSIGHTED = "Nearsightedness",
  NEBmen = "Number of Children (men)",
  NEBwomen = "Number of Children (women)",
  NEURO = "Neuroticism",
  OPEN = "Openness",
  READING = "Age Started Reading",
  RELIGATT = "Religious Attendance",
  RISK = "Risk Tolerance",
  SELFHEALTH = "Self-Rated Health",
  SELFMATH = "Self-Rated Math Ability",
  SWB = "Subjective Well-Being",
  VOICEDEEP = "Age Voice Deepened",
  WORKSAT = "Satisfaction with Work"
)

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants. PGS are rescaled to zero mean and unit variance. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data. For our purposes, this is equivalent to residualizing on principal components [@wurm2014residualizing]. Effectively, our analysis is of the residualized PGS that are widely used in practical applications. \[TODO actually we could just residualize; only the SE of the PCs is affected\]

@@

The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection. However the approach is also based in quantitative genetic theory and is able to yield an approximate estimate of genetic change from one generation to the next.

The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations. In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness. As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score from one generation. We standardize our polygenic scores to have a mean of 0 and a standard deviation of 1. With this standardization the regression coefficient of RLRS on the PGS is equal to the covariance. Our approach to measuring natural selection is not novel, having been and detailed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].

@@

# The large n_reps is to get an accurate upper tail on the black sample, 
# since it is close to 0:
n_reps <-  30 #599

@@

if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
if (significant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$black significant")
if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
if (insignificant(ci_boot_gkids$white)) warning("ci_boot_gkids$white insignificant")
if (significant(ci_boot_gkids$black)) warning("ci_boot_gkids$black significant")
```

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Controls include 10 principal components of genetic array
#|   data. Dashed lines are fitted from linear regressions.

@@

The HRS has data on the years of education for the parents of respondents. We recalculate the coefficients on education using the parent's PGS, in case the genetics of education and human capital changed between generations. We then correlate these coefficients with the number of siblings, finding a significant result (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`). The correlation is only slightly smaller and the confidence intervals are only slightly wider. This is surprising since parent's education is likely recorded with error and the respondent's polygenic score is only a proxy for their parents.

@@

The EA polygenic score captures a key element of human capital and was trained on a GWAS sample size of one million, making it a sensible PGS with which to test the economic theory. In TABLE XXX we report the main effect of the EA polygenic score and its interaction effect on RLRS, for each of the four putative moderators. Only for the case of income is the interaction effect significant at p \< 0.05. We might wonder whether the weak effects are because of insufficient power or because any real interaction is small in the sample. In @hugh2022human the interaction effects were large, often as big as the main effects. Yet in the HRS, the point estimates of the interactions appear generally smaller, suggesting the nulls may not just be a consequence of lower power.

We should attain more statistical power if we utilise information from other polygenic scores at the same time. However, some caution is necessary. Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate. To perform, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, 
               position = position_jitter(width = 0.1, seed = 12345)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    geom_text(aes(label = outliers), size = 2, colour = "black", 
              position = position_jitter(width = 0.1, seed = 12345), 
              hjust = -0.15) +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

Lastly, we would like to know natural selection's effect sizes. The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations. Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS. They can be scaled up by

@@

Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century. Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed. The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting. The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.

<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->

@@

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The result is consistent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929. The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is similar to previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.

@@

There are `r n_black` genotyped black survey participants. Among this sample, the relationship between polygenic scores' coefficients on RLRS and on education is negative but insignificant and imprecisely estimated (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Correlations between number of siblings and education are insignificant (correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). Similarly, coefficients on RLRS2 (children's average number of children) were insignificantly correlated with coefficients on years of education, with large standard errors (correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`; removed `r gkids_unusable_black` respondents reporting grandchildren but no children).

@@

```{r}
#| label: fig-rlrs
#| fig-cap: |
#|  Coefficients of PGS on RLRS, controlling for 10 principal components of
#|  genomic array data. Lines are 95% confidence intervals.
#| fig-height: 6

@@

```{r}
#| label: fig-siblings
#| fig-cap: |
#|   Scatterplot of PGS coefficients on number of live siblings and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.

@@

```{r}
#| label: fig-gkids
#| fig-cap: |
#|   Scatterplot of PGS coefficients on RLRS2 and years of
#|   education. RLRS2 is the reproductive success of respondent's offspring
#|   relative to the success of other offspring. Controls include 10 principal
#|   components of genetic array data. Dashed lines show linear regressions.

>>>>

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: scores which correlate negatively (positively) with
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  half borne out: coefficients are larger only among low-income
  parents and unmarried parents. Using respondents' number of 
  siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

knitr::knit_hooks$set("inline", 
  \(x) {
    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
    }
    x
  }
)
nice_names <- c(
  SCZ = "Schizophrenia",
  WC = "Waist circ.",
  WHR = "Waist-hip ratio",
  NEUROT = "Neuroticism",
  WELLB = "Well-being",
  DEPSYMP = "Depr. symptoms",
  CD = "Cor. art. disease",
  MI = "Myoc. infarction",
  CORT = "Cortisol",
  T2D = "Type 2 diabetes",
  BIP = "Bipolar",
  ADHD_PGC17 = "ADHD",
  XDISORDER = "Cross-disorder",
  MENA = "Age at menarche",
  MENO = "Age at menopause",
  EXTRAVER = "Extraversion",
  AUTISM = "Autism",
  AB = "Antisoc. behav.",
  OCD = "OCD",
  AFBC = "Age 1st birth combn",
  AFBF = "Age 1st birth F",
  AFBM = "Age 1st birth M",
  MDD2 = "Maj. depr. disorder",
  PTSDAA = "PTSD Afr",
  PTSDEA = "PTSD Eur",
  PTSDC = "PTSD combined",
  HDL = "HDL cholesterol",
  LDL = "LDL cholesterol",
  TC = "Total cholesterol",
  ANXFS = "Anxiety factor",
  ANXCC = "Anxiety case-control",
  BUN = "Blood urea nitro.",
  BUNTE = "Blood urea nitro. TE",
  CKD = "Kidney disease",
  CKDTE = "Kidney disease TE",
  DBP = "Diastolic blood pr.",
  BMI2 = "BMI",
  HEIGHT2 = "Height",
  AI = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW = "Drinks per week",
  SC = "Smoking cessation",
  SI = "Smoking initiation",
  HTN = "Hypertension",
  CANNABIS = "Cannabis",
  GWAD2NA = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC = "Alcohol dependence",
  PP = "Pulse pressure",
  SBP = "Systolic blood pr.",
  EGFR = "Glom. filtr.",
  EGFRTE = "Glom. filtr. TE",
  EA3 = "Educ. attainment",
  HBA1CAA = "HbA1c Afr",
  HBA1CEA = "HbA1c Eur",
  GCOG2 = "Gen. cognition",
  ACTIVITY = "Physical Activity", # pgirep here
  ADHD = "ADHD",
  ADVENTURE = "Adventurous",
  AFB = "Age First Birth",
  ALLERGYCAT = "Cat Allergy",
  ALLERGYDUST = "Dust Allergy",
  ALLERGYPOLLEN = "Pollen Allergy",
  ASTECZRHI = "Asthma/Eczema/Rhinitis",
  ASTHMA = "Asthma",
  AUDIT = "Alcohol Misuse",
  BMI = "Body Mass Index",
  CANNABISpgirep = "Cannabis Use",
  COGEMP = "Cognitive Empathy",
  COPD = "Chronic Obstructive Pulmonary Disease",
  CPD = "Cigarettes per Day",
  CP = "Cognitive Performance",
  DELAYDISC = "Delay Discounting",
  DEP = "Depressive Symptoms",
  DPWpgirep = "Drinks per Week",
  EA = "Educational Attainment",
  EVERSMOKE = "Ever Smoker",
  EXTRA = "Extraversion",
  FAMSAT = "Satisfaction with Family",
  FINSAT = "Satisfaction with Finances",
  FRIENDSAT = "Satisfaction with Friends",
  HAYFEVER = "Hayfever",
  HEIGHTpgirep = "Height",
  HIGHMATH = "Highest Math",
  LEFTOUT = "Left Out of Social Activity",
  LONELY = "Loneliness",
  MENARCHE = "Age First Menses",
  MIGRAINE = "Migraine",
  MORNING = "Morning Person",
  NARCIS = "Narcissism",
  NEARSIGHTED = "Nearsightedness",
  NEBmen = "Number of Children (men)",
  NEBwomen = "Number of Children (women)",
  NEURO = "Neuroticism",
  OPEN = "Openness",
  READING = "Age Started Reading",
  RELIGATT = "Religious Attendance",
  RISK = "Risk Tolerance",
  SELFHEALTH = "Self-Rated Health",
  SELFMATH = "Self-Rated Math Ability",
  SWB = "Subjective Well-Being",
  VOICEDEEP = "Age Voice Deepened",
  WORKSAT = "Satisfaction with Work"
)

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants. PGS were residualised on the first ten within-ethnicity principal components of the DNA array data, to reduce bias from population stratification. PGS were then rescaled to zero mean and unit variance.

@@

The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection. However the approach is also based on quantitative genetic theory and is able to yield an estimate of genetic change from one generation to the next.

The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations. In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness. As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score per generation. Because the polygenic scores are standardized to have a variance of 1, the regression coefficient of RLRS on the PGS is equal to the covariance. Our approach to measuring natural selection is not novel, having been employed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].

@@

# The large n_reps is to get an accurate upper tail on the black sample, 
# since it is close to 0:
n_reps <-  599

@@

if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
if (insignificant(ci_boot_fert$black)) warning("ci_boot_fert$black insignificant")
if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$black significant")
if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
if (insignificant(ci_boot_gkids$white)) warning("ci_boot_gkids$white insignificant")
if (significant(ci_boot_gkids$black)) warning("ci_boot_gkids$black significant")
```

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Dashed lines are fitted from linear regressions.

@@

The HRS has data on the years of education for the parents of respondents. We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations. We then correlate these new coefficients with the number of siblings, finding a positive correlation again (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`). The estimate is slightly smaller as would be expected since respondents likely recall their parent's education with some error.

@@

We should attain more statistical power if we utilise information from other polygenic scores at the same time. However, some caution is necessary. Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate. We test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, 
               position = position_jitter(width = 0.125, height=0.001, seed = 12346)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    geom_text(aes(label = outliers), size = 2, colour = "black", 
              position = position_jitter(width = 0.125,  height=0.001, seed = 12346), 
              hjust = -0.15) +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

Lastly, we would like to know natural selection's effect sizes. The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations. Polygenic scores contain error, so estimated effects are biased towards zero compared to the effect of the true PGS. They can be scaled up by

@@

Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century. Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased in the late twentieth century, contrary to how we believe the genetics have changed. However, the prevalence of self-reported anxiety and depressive symptoms have increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism. The prevalence of asthma has also increased [@Moorman2007] in line with the polygenic score. The apparent paradox of phenotypes and nautral selection moving in opposite directions suggests that environmental change across history is a much greater force than that of natural selection.

@@

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is always more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The result is congruent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929. The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is slightly smaller than in previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.

@@

There are `r n_black` genotyped black survey participants. Among this sample, the relationship between polygenic scores' coefficients on RLRS and on education is negative and significant (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Correlations between number of siblings and education are insignificant (correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). Similarly, coefficients on RLRS2 (children's average number of children) were insignificantly correlated with coefficients on years of education, with large standard errors (correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`; removed `r gkids_unusable_black` respondents reporting grandchildren but no children).

@@

```{r}
#| label: fig-rlrs
#| fig-cap: |
#|  Coefficients of PGS on RLRS. Lines are 95% confidence intervals.
#| fig-height: 6

@@

```{r}
#| label: fig-siblings
#| fig-cap: |
#|   Scatterplot of PGS coefficients on number of live siblings and years of
#|   education. Dashed lines show linear regressions.

@@

```{r}
#| label: fig-gkids
#| fig-cap: |
#|   Scatterplot of PGS coefficients on RLRS2 and years of
#|   education. RLRS2 is the reproductive success of respondent's offspring
#|   relative to the success of other offspring. Dashed lines show linear regressions.
